Developmental stage determines the realized energy landscape for a flight specialist

Version 1

Abstract

Abstract

The heterogeneity of the physical environment determines the cost of transport for animals, shaping their energy landscape. Animals respond to this energy landscape by adjusting their distribution and movement to maximize gains and reduce movement costs. Much of our current knowledge about energy landscape dynamics focuses on factors external to the animal, particularly the spatio-temporal variations of the environment. However, an animal’s internal state can significantly impact its ability to perceive and utilize the available energy, creating a distinction between the “fundamental” and the “realized” energy landscapes. Here we show that the realized energy landscape varies along the onto-genetic axis. Locomotor and cognitive capabilities of individuals change over time, especially during the early life stages. We investigate the development of the realized energy landscape in the Central European Alpine population of the golden eagle Aquila chrysaetos, a large predator that requires negotiating the atmospheric environment to achieve energy-efficient soaring flight. We quantified weekly energy landscapes using environmental features for 55 juvenile golden eagles, demonstrating that energetic costs of traversing the landscape decreased as the birds aged. In fact, the potentially flyable area within the Alpine region increased 2,170-fold during their first three years of independence. Our work contributes to achieving a predictive understanding of animal movement behaviors by presenting ontogeny as a mechanism shaping the realized energy landscape.

Introduction

The physical characteristics of the environment determine the energetic cost of transport for animals, defining their “energy landscape” [1, 2]. Consequently, temporal dynamics in environmental conditions can affect the inherently spatially-explicit energy landscapes across time. For example, patterns of thawing and freezing of lakes and rivers shape the migratory routes of Caribou, as walking on ice is energetically preferable to swimming [3]. Variations of the energy landscape along the temporal axis are most notable for animals that move in moving media: water or air. Sharks adjust their core habitat use to make the most of temporary updrafts formed by tidal currents colliding with upward-facing slopes, which help them reduce the energetic costs of remaining buoyant [4]. Temporal variations in atmospheric conditions shape hourly [2] and seasonally [5] optimal flight routes for birds.

The spatio-temporal variation in the energy landscape is predictable to a great extent. Animals reduce and thus optimize the cost of transport by tracking these changes as they move through the energy landscapes. As such, the distribution and movement patterns of animals can be largely explained by the energy landscape [6], making energy landscapes useful for understanding ecological and evolutionary processes. This knowledge is increasingly important for policy making to reduce human-wildlife conflicts, particularly for development of infrastructures [7, 8], and more generally in conservation planning [9]. Based on the timing and location of movement the energy landscape can be specific yet predictable for animals with similar morphology and movement modes, at the guild [2, 10], species [8], or sub-species [11] levels.

To date, our understanding of animals’ interaction with the energy landscape has depended on factors extrinsic to the moving animal. Reconstructing the energy landscape solely based on the characteristics of the environment can be conceptualized similar to Hutchinson’s concept of the fundamental ecological niche [12, 13]. Hutchinson defined the fundamental niche as the full range of environmental conditions in which a species can potentially exist and reproduce. Here we argue that the energy landscape too can, or possibly should, be conceptualized as the fundamental movement niche, defining the full range of physical environmental conditions within which a species can move while adhering to its movement-related energy budget, or sustainable cost of transport. However, just as animals do not occupy the entirety of their fundamental Hutchinsonian niche in reality [13], usually due to competition, various factors can contribute to an animal not having access to the entirety of its fundamental movement niche. Our current understanding of the energy landscape lacks a differentiation between a fundamental and a realized energy landscape.

An animal’s realized energy landscape should depend on the motive to move and its internal state. For example, an animal’s decision-making when moving might be the outcome of its strategy to maximize long-term fitness rather than short-term energy savings. Halsey et al. [14] introduced the concept of “individual energy landscapes”, stating that the energy landscape fluctuates, for example, when an animal increases its speed to out-run a predator [14] or when infection or illness reduces an animal’s movement capacity [15, 16]. As a consequence, the realized energy landscape varies in time from the perspective of the moving animal, sometimes irrespective of, but in interaction with and as a consequence of, the immediate environment.

We would like to argue that the animals’ ability to exploit the energy landscape is transformed across one of the most fundamental progressions in all animals’ lives: the ontogenetic axis. As young animals progress through their developmental stages, their movement proficiency [17] and cognitive capabilities [18] improve and memory manifests [19]. Cognitive capabilities allow the animal to improve the perception of its environment and to make optimal decisions when choosing energy-efficient paths [20, 21]. Improved motor skills enable the animal to respond more appropriately to the perceived environment to utilize the maximum available energy and/or to avoid areas expensive to traverse [22–25]. And finally, memory allows the animal to recall past combinations of space and environmental conditions that affected its costs of movement [26]. Developmental changes and improvements in all these elements should be reflected in the animal’s realized energy landscape, where hotspots of energy gain within the landscape expand as young animals gain experience and master their specific movement behaviors.

We investigated the demographic shift in the realized energy landscape through gaining experience in a movement specialist, the golden eagle Aquila chrysaetos. As obligate soaring birds, golden eagles utilize ascending air currents, known as uplifts, to gain vertical altitude, which subsequently enables them to glide for horizontal displacement, without relying on energetically expensive wing flapping. Soaring flight is a learned and acquired behavior [24, 27], requiring advanced cognitive skills to locate uplifts as well as fine-tuned locomotor skills for optimal adjustment of the body and wings to extract the most energy from them. This flight mode enables soaring birds to cover long distances with minimal energy investments, almost as low as when resting [28]. We approach the energy landscape concept with a focus on energy availability. This is commonly done for soaring birds, as uplift and wind support can directly provide energy for flying birds [11, 29].

We used 46,000 hours of flight data collected using bio-logging devices during the transience phase of natal dispersal (hereafter post-emigration) of 55 wild-ranging golden eagles in the Central European Alps to test our hypothesis that the realized energy landscape of young birds gradually expands as they age. More specifically, we expected the habitat to become progressively cheaper to traverse as the birds aged, resulting in an expansion of flyable areas within the Alpine region as a consequence of gaining experience and flight proficiency.

Results

We explored the movement decisions of 55 juvenile golden eagles in the Central European Alps during their post-emigration commuting flights (i.e. non-stop flight bouts lasting at least 1 hour) from the first week until three years after emigration. Following a step-selection approach30, at each movement step, we compared the observed step to 50 alternative steps that were available to each bird in space and time. These steps were compared in respect to the topography of the terrain, specifically Topographic Ruggedness Index (TRI) and distance to ridge lines, both useful proxies for occurrence of uplifts29,31. The resulting step-selection model distinguishing between the used and available conditions with a good predictive performance based on the normalized Root Mean Squared Error (RMSE = 0.14). Overall, the birds preferred to fly over areas with high potential for uplift formation, characterized by high values of TRI and lower distance to ridge lines (Fig. 1). The birds also preferred to fly with long step lengths, a trend that increased with age. As the birds grew older, they were less likely to avoid areas with lower TRI and higher distance to ridge lines (Fig. 2). We found individual variation in the coefficients for TRI and distance to ridge line (Fig. S1).

The coefficient estimates of the step selection function fitted using glmmTMB. All variables were z-transformed prior to modeling. The error bars show 95% confidence intervals.

Flyability index predicted using the step-selection model for combinations of topography and week since emigration values. The interactions between Topographic Ruggedness Index (TRI) and distance to ridge lines with week are among the set of criteria that young eagles used for selecting where to fly during their commuting flights.

We used the step-selection model to predict the flyability of the Alpine region for each developmental stage (week since emigration). This flyability index quantified the potential energy that a bird could obtain from its surroundings by considering topography and the bird’s movement. This index serves as a measure of the suitability of a location for energy-efficient flight, reflecting the bird’s ability to exploit favorable uplift conditions. We constructed energy landscapes for the juvenile golden eagles for each week since independence by predicting the flyability within the Alpine region. Based on these predictions, we determined the flyable area for the juvenile and immature golden eagles during the developmental phase at weekly increments. We defined flyable areas as the cells within a 100*100 m grid with a flyability value larger than 0.7. Our analysis show that hotspots of energy availability in the birds’ landscape expanded over time (Fig. 3; Video S1). We found that the flyable area increased 2,170-fold from the first week until three years after emigration. From an initial 0.038 % in the first week after emigration, the flyable area followed a logistic growth curve to plateau at 81 % of the entire Alpine region becoming flyable in the third year (Fig. 4). The tracking data indicated that the golden eagles flew over an area of 92,000 km2 by the third year after emigration.

Hotspots of energy availability for golden eagles’ flight in the Alps. Flyable areas were defined as cells within a 100*100 m grid with predicted flyability above 0.7 based on our step-selection model. The maps show the 2D kernel density estimation of flyable areas for golden eagles at different timestamps since dispersal: week 1, week 4 (one month), week 24 (six months) and week 52 (one year). The raw prediction maps for every week since dispersal are shown in the video S1.

The flyable area for juvenile golden eagles in the Alpine region from the first week until three years after emigration. Flyable area was defined as the total number of cells within a 100*100 m grid with predicted flyability larger than 0.7 based on the step-selection model. The positive trend shows that juvenile golden eagles can fly over a larger portion of the Alpine region as they age.

Discussion

We show that the flyable area for juvenile golden eagles increased during the developmental period. This increase, we argue, is most likely linked to the birds’ improving ability to negotiate their atmospheric environment better as they aged, due to an increasing capacity to perceive and exploit the energy available in the landscape. Because we estimated flyability based on proxies of uplift formation, an important component shaping the energy availability landscape for soaring birds [11, 29], the gradual increase in flyable areas can be interpreted as a change in the realized energy landscape.

Our findings assert the concept of “individual energy landscapes”. Initially described by Halsey14, this concept suggests that the energy expenditure of an animal when moving is more complex than the cost of transport as determined solely by its morphology and the external environment. The locomotor decisions that an animal makes to maximize its fitness at any given point in time would also affect its energy expenditure. We present evidence that ontogeny also serves as a potential mechanism giving rise to individual energy landscapes, not only due to its impact on decision-making, but also due to the influence it exerts on the available options for the animal. During the early stages of their development, animals may not possess the full capacity to perceive their environment, as they will in later stages [20, 21]. Additionally, their motor skills are still being honed24. Consequently, younger individuals may respond less efficiently to the environmental conditions they must navigate, in contrast to older, more proficient individuals.

By relying on static topographic variables to build the energy landscapes, we ensured that the environment remained constant over time. Given that golden eagles’ wing morphology remains unchanged during the post-emigration period from their natal territory32,33, the changes in the energy availability for the birds represented their mastering of soaring flight. Our findings reveal that as the eagles aged, they adopted longer step lengths, indicating an increasing ability to sustain longer uninterrupted flight bouts.

Furthermore, our results provide insight into the development of different flight modes. These birds can switch between orographic and thermal soaring34. Orographic soaring refers to utilization of upward air currents generated when wind encounters and flows over elevated topographic features such as hills, mountains, or cliffs. Locating orographic uplifts requires less experience, as they form predictably at mountain ridges35,36, compared to locating and using more transient thermals which form less predictably over flat surfaces29. Our findings suggest that with increasing age, golden eagles place decreasing importance on ruggedness and proximity to ridge lines when deciding where to fly, indicating a learning progression from mastering orographic uplifts before tackling the more challenging thermal soaring.

The energy landscapes maps showed the amount of energy potentially available to juvenile golden eagles during progressing weeks since emigration (Video S1). These maps are based on a step-selection model built on the birds’ empirical use of the landscape, the energy availability thus represents the energy that would be theoretically perceivable and exploitable by the birds in a specific developmental phase or age. The detailed perspective on an energy landscape which transforms over time with the accumulation of experience highlights the process of changes in the realized movement niche that would be available to the animals at a certain age. These changes in the ability to extract energy from the atmospheric column will induce age-specific consequences on movement capacity and the cost of transport.

We argue that the realized energy landscape does not represent the realized ecological niche nor the distribution range of animals, but rather the subset of the fundamental energy landscape that is available to the individual based on its ability to perceive and exploit the energy in the landscape. In our study, the realized movement patterns of the golden eagles within the Alpine region covered a small portion of the area predicted as flyable. This is because the birds’ movement (and thus their realized individual and age-specific movement niches) will be defined not only by the energy landscape, but also by the interplay of philopatry, developmental requirements, the social landscape, and anthropogenic features. As golden eagles age, other crucial behaviors such as transitioning from predominantly scavenging to active hunting are also happening37. A transition that in itself also might depend on advanced flying skills. Thus, the choice of where to fly will not solely follow an energy-minimizing strategy while moving, but naturally also increasingly accommodate resource use opportunities. Additionally, older transient golden eagles are more susceptible to potentially deadly conflicts with territorial adult birds, influencing their movement decisions as they avoid established territories38 affecting their landscape of fear and thus their decision landscape overall39–41. The distribution of anthropogenic features could also impact the eagles’ decision-making on where to fly42. The strong individual variation that we observed in the birds’ response to distance to ridge lines (Fig. S1) could indicate variable landscape utilization strategies among the juvenile golden eagles in response to these factors and their interaction.

Conclusion

Spatial maps serve as valuable tools in informing conservation and management strategies by showing the general distribution and movement patterns of animals. Within this context, energy landscapes play an important role in identifying potential areas of conflict between animals and anthropogenic infrastructures such as wind farms. The predictability of environmental factors that shape the energy landscape has facilitated the development of these conservation tools, which have been extrapolated to animals belonging to the same ecological guild traversing similar environments. However, with recent development of concepts of species-specific8 and individual-specific energy landscapes14, the realistic prediction of animals’ energy landscapes appears a multifaceted challenge. By identifying the distinct dimensions along which individual energy landscapes can vary, particularly along the demographic axis, we enhance our capacity to understand and predict these landscapes. Our research contributes to this broader understanding by revealing the stark influence of ontogeny in defining the aging of the realized energy landscape. By taking inspiration from the ecological niche definitions to conceptualize the energy landscape, we hope to improve further development of this concept, aiding the interpretation of variations in the energy landscape across many axes, both external and internal to the moving animal.

Materials and Methods

Bio-logging dataset

We used Global Positioning System tracking data from a long-term study of juvenile golden eagles in the resident population of Central European Alps. We used data collected for a total of 55 juveniles tracked over 2017-2023 (Table S1) with a regular sampling schedule of one GPS fix every 20 minutes during daytime (see43 for details on fieldwork procedure and data sampling schedules). Nestlings were fitted with solar powered loggers manufactured by e-obs GmbH (Munich, Germany) using a leg-loop harness. The weight of the harness and logger (in total maximum 60g) was below the recommended maximum 3% of the birds’ body mass. We subset the data to correspond to the transient phase of natal dispersal up to three years after emigration. We identified emigration timing using recursions to the natal territory44. If an animal spent more than 14 consecutive days without recursions to a 7 km radius around its nest, we classified it as dispersed. These thresholds were chosen based on expert knowledge of golden eagle behavior in the region. For two animals that had exceptionally large natal territories to start with, we increased the radii to 30 km to compensate for this potential bias.

Track segmentation

We focused on commuting flights for investigating the energy landscape for the golden eagles. To extract commuting flights, we segmented the tracking data based on ground speed and flight altitude using Expectation-Maximization binary Clustering (EMbC)45 applied at the population level (i.e. to all individuals). To calculate flight altitude above ground for each tracking point, we first calculated altitude above mean sea level as the difference between altitude above ellipsoid recorded by the bio-logging devices and the Global Geoid Model EGM96 (WGS 84). We then subtracted the ground elevation (Copernicus Digital Elevation Model; 25 m resolution https://land.copernicus.eu/) from altitude above mean sea level as an estimate of flight altitude. We labeled segments with high flight altitude and high speed as commuting flights and used these for further analysis. We sub-sampled the tracking data to hourly intervals to focus on commuting flights that were longer than one hour.

Topographic variables

We used two topographic variables as proxies for uplift availability: Topographic Ruggedness Index (TRI) and distance to ridge lines. The former describes the topographic heterogeneity of the terrain by calculating the mean of the absolute elevation difference between a cell and its adjacent cells46 and the latter represents a proxy for orographic updrafts generated by deflection of the air current by a physical barrier31,35. To extract the ridge lines and TRI values, we used the Copernicus Digital Elevation Model. We identified the ridge lines based on the Weisss landforms classification using Topographic Position Index (TPI), a scale-dependent index which compares the elevation of each cell to the neighboring cells within a given radius47. As the landform categories depend on the size of the radius, several TPI classifications at different scales were obtained using different radii48,49. We selected a classification using a 10 m inner radius and 200 m outer radius on the basis of the narrow size of the identified ridge lines. Then, we calculated the distance from each pixel to the ridge line pixel using the function proximity raster from SAGA50 to obtain a distance map to the ridge line. For both TRI and distance to ridge lines, we averaged neighboring cell values to achieve a 100 m cell size.

Modeling the energy landscape

We used a step-selection approach30 to summarize the environmental conditions in which the birds chose to commute. Each pair of consecutive points in the tracking dataset was considered as one step. For each step, we generated 50 random steps that originated at the observed steps’ start point, but had alternative end points to represent alternative movement decisions that the bird could have taken. The location of the alternative end points were selected by drawing from a gamma distribution fitted to the step lengths and a von Mises distribution fitted to the turning angles of all the empirical tracking data. The result was a stratified dataset with one used and 50 alternative points in each stratum.

All used and available points were associated with TRI and distance to ridge line values. We found no autocorrelation (r < 0.5) among our predictor variables TRI, distance to ridge line, step length, and week since emigration (as a continuous variable). The predictor variables were z-transformed prior to modeling to ensure comparability of the coefficient estimates. We used a conditional logistic regression to build a step-selection function to model and predict the energy landscape. We included step length as one of the predictors in our model, to take the movement characteristics of the growing juveniles into account30. To track the changes in the importance of the topographic variables over time, we included three-way interaction terms for the TRI and distance to ridge line, with step length and week since emigration. To account for individual variation, we included the random effect of individuals on the slopes of TRI and distance to ridge line. The model was fit using the Template Model Builder (TMB;51) in R52, following the procedure suggested by Muff et al.53.

To test the predictive performance of the model, we computed the normalized Root Mean Squared Error (RMSE)54. This metric can be interpreted as the standard deviation of the unexplained variance, with lower values indicating a smaller difference between observed data and the model’s predicted values.

To better interpret the interaction terms between the topographic variables and week since emigration, we made predictions using the model for different combinations of TRI and week as well as distance to ridge line and week. In the new datasets that we generated to make these predictions, we set the variables that were not represented in the interaction term to their average values and set the grouping factor that represented the stratum as missing data. We made the predictions on the scale of the link function and converted them to values between 0 and 1 using the inverse logit function55. We considered these predicted values as the flyability index, representing the potential energy available in the landscape to support flight, based on the uplift proxies (TRI and distance to ridge line) and the movement capacity (step length) of the birds included in the model.

Energy landscape maps

To investigate the changes in the energy landscape across the ontogenetic axis, we used our model to create prediction maps for weekly increments during the first three years after emigration. We made our prediction maps for the entire extent of the Alpine region (“Perimeter of the Alpine Convention” layer available via Permanent Secretariat of the Alpine Convention). For each week since emigration, we created a new dataset to make predictions for each 100 m x 100 m cell. We assigned the mean step length value for each given week in the new datasets.

To explore the changes in flyable areas over time, we defined landscape flyability as flyability values above 0.7. We then estimated the total flyable areas for each week since emigration. The Alpine region covers 190,544.7 km2 (based on the Alpine perimeter layer mentioned above). As a final step, we compared the eagles’ flyable area in weeks 1 and 156 with the total area of the Alpine region.

Supporting information

eLife Assessment

This important study substantially advances our understanding of energy landscapes and their link to animal ontogeny. The evidence supporting the conclusions is compelling, with high-throughput telemetry data and advanced track segmentation methods used to develop and map energy landscapes. The work will be of broad interest to animal ecologists.

Reviewer #1 (Public Review):

Summary:

The authors propose that the energy landscape of animals can be thought of in the same way as the fundamental versus realized niche concept in ecology. Namely, animals will use a subset of the fundamental energy landscape due to a variety of factors. The authors then show that the realized energy landscape of eagles increases with age as the animals are better able to use the energy landscape.

Strengths:

This is a very interesting idea and that adds significantly to the energy landscape framework. They provide convincing evidence that the available regions used by birds increase with size.

Weaknesses:

Some of the measures used in the manuscript are difficult to follow and there is no mention of the morphometrics of birds or how these change with age (other than that they don't change which seems odd as surely they grow). Also, there may need to be more discussion of other ontogenetic changes such as foraging strategies, home range size etc.

Reviewer #2 (Public Review):

Summary:

With this work, the authors tried to expand and integrate the concept of realized niche in the context of movement ecology by using fine-scale GPS data of 55 juvenile Golden eagles in the Alps. Authors found that ontogenic changes influence the percentage of area flyable to the eagles as individuals exploit better geographic uplifts that allow them to reduce the cost of transport.

Strengths:

Authors made insightful work linking changes in ontogeny and energy landscapes in large soaring birds. It may not only advance the understanding of how changes in the life cycle affect the exploitability of aerial space but also offer valuable tools for the management and conservation of large soaring species in the changing world.

Weaknesses:

Future research may test the applicability of the present work by including more individuals and/or other species from other study areas.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer 1 (Public Review):

Summary:

The authors propose that the energy landscape of animals can be thought of in the same way as the fundamental versus realized niche concept in ecology. Namely, animals will use a subset of the fundamental energy landscape due to a variety of factors. The authors then show that the realized energy landscape of eagles increases with age as the animals are better able to use the energy landscape. Strengths:

This is a very interesting idea and that adds significantly to the energy landscape framework. They provide convincing evidence that the available regions used by birds increase with size.

Weaknesses:

Some of the measures used in the manuscript are difficult to follow and there is no mention of the morphometrics of birds or how these change with age (other than that they don’t change which seems odd as surely they grow). Also, there may need to be more discussion of other ontogenetic changes such as foraging strategies, home range size etc.

We thank reviewer 1 for their interest in our study and for their constructive recommendations. We have included further discussions of these points in the manuscript and outline these changes in our responses to the detailed recommendations below.

Reviewer 2 (Public Review):

Summary:

With this work, the authors tried to expand and integrate the concept of realized niche in the context of movement ecology by using fine-scale GPS data of 55 juvenile Golden eagles in the Alps. Authors found that ontogenic changes influence the percentage of area flyable to the eagles as individuals exploit better geographic uplifts that allow them to reduce the cost of transport.

Strengths:

Authors made insightful work linking changes in ontogeny and energy landscapes in large soaring birds. It may not only advance the understanding of how changes in the life cycle affect the exploitability of aerial space but also offer valuable tools for the management and conservation of large soaring species in the changing world.

Weaknesses:

Future research may test the applicability of the present work by including more individuals and/or other species from other study areas.

We are thankful to reviewer 2 for their encouragement and positive assessment of our work. We have addressed their specific recommendations below.

Recommendations for the authors:

Reviewer 1 (Recommendations For The Authors):

I found this to be a very interesting paper which adds some great concepts and ideas to the energy landscape framework. The paper is also concise and well-written. While I am enthusiastic about the paper there are areas that need clarifying or need to be made clearer. Specific comments below:

Line 64: I disagree that competition is the fundamental driver of the realized niche. In some cases, it may be but in others, predation may be far more important (as an example).

We agree with this point and have now clarified that competition is an example of a driver of the realized niche. We have also included predation as another example:

"However, just as animals do not occupy the entirety of their fundamental Hutchinsonian niche in reality [1], for example due to competition or predation risk, various factors can contribute to an animal not having access to the entirety of its fundamental movement niche."

Intro: I think the authors should emphasize that morphological changes with ontogeny will change the energy landscape for many animals. It may not be the case specifically with eagles but that won’t be true for other animals. For example, in many sharks, buoyancy increases with age.

We agree and have now clarified that the developmental processes that we are interested in happen in addition to morphological changes:

"In addition to morphological changes, as young animals progress through their developmental stages, their movement proficiency [2] and cognitive capabilities [3] improve and memory manifests [4]."

Line 91-93: The idea that birds fine-tune motor performance to take advantage of updrafts is a very important one to the manuscript and should be discussed in a bit more detail. How? At the moment there is a single sentence and it doesn’t even have a citation yet this is the main crux of the changes in realized energy landscape with age. This point should be emphasized because, by the end of the introduction, it is not clear to me why the landscape should be cheaper as the birds age?

Thank you for pointing out this missing information. We have now added examples to clarify how soaring birds fine-tune their motor performance when soaring. These include for example adopting high bank angles in narrow and weak thermals [5] and reducing gliding airspeed when the next thermal has not been detected [6]:

"Soaring flight is a learned and acquired behavior [7, 8], requiring advanced cognitive skills to locate uplifts as well as fine-tuned locomotor skills for optimal adjustment of the body and wings to extract the most energy from them, for example by adopting high bank angles in narrow and weak thermals [5] and reducing gliding airspeed when the next thermal has not been detected [6]."

Results:

Line 106: explain the basics of the life history of the birds in the introduction. I have no idea what emigration refers to or the life history of these animals.

Thank you for pointing out the missing background information. We have now added this

information to the introduction:

"We analyzed 46,000 hours of flight data collected from bio-logging devices attached to 55 wild-ranging golden eagles in the Central European Alps. These data covered the transience phase of natal dispersal (hereafter post-emigration). In this population, juveniles typically achieve independence by emigrating from the parental territory within 4-10 months after fledging. However, due to the high density of eagles and consequently the scarcity of available territories, the transience phase between emigration and settling by eventually winning over a territory is exceptionally long at well over 4 years. Our hypothesis posited that the realized energy landscape during this transience phase gradually expands as the birds age."

What I still am having a hard time understanding is the flyability index. Is this just a measure of the area animals actively select and then the assumption that it’s a good region to fly within?

We have modified our description of the flyability index for more clarity. In short, we built a step-selection model and made predictions using this model. The predictions estimate the probability of use of an area based on the predictors of the model. For the purpose of our study and what our predictors were (proxies for uplift + movement capacity), we interpreted the predicted values as the "flyability index". We have now clarified this in the methods section:

"We made the predictions on the scale of the link function and converted them to values between 0 and 1 using the inverse logit function [9]. These predicted values estimated the probability of use of an area for flying based on the model. We interpreted these predicted values as the flyability index, representing the potential energy available in the landscape to support flight, based on the uplift proxies (TRI and distance to ridge line) and the movement capacity (step length) of the birds included in the model."

It might also be useful to simply show the changes in the area the animals use with age as well (i.e. a simple utilization distribution). This should increase in age for many animals but would also be a reflection of the resources animals need to acquire as they get older.

We have now added the figure S2 to the supplementary material. This plot was created by calculating the cumulative area used by the birds in each week after emigration. This was done by extracting the commuting flights for each week, converting these to line objects, overlapping the lines with a raster of 100*100 m cell size, counting the number of overlapping cells and calculating the area that they covered. We did not calculate UDs or MCPs because the eagles seem to be responding to linear features of the landscape, e.g. preferring ridgelines and avoiding valleys. Using polygons to estimate used areas would have made it difficult to ensure that decision-making with regards to these linear features was captured.

In a follow-up project, a PhD student in the golden eagle consortium is exploring the individuals’ space use after emigration considering different environmental and social factors. The outcome of that study will further complete our understanding of the post-emigration behavior of juvenile golden eagles in the Alps.

How much do the birds change in size over the ontogeny measured? This is never discussed.

Thank you for bringing up this question. The morphometrics of juvenile golden eagles are not significantly different from the adults, except in the size of culmen and claws [10]. Body mass changes after fledging, because of the development of the pectoral muscles as the birds start flying. Golden eagles typically achieve adult-like size and mass within their natal territory before emigration, at which time we started quantifying the changes in energy landscape. Given our focus on post-emigration flight behavior, we do not expect any significant changes in size and body mass during our study period. We now cover this in the discussion:

"Juvenile golden eagles complete their morphological development before gaining independence from their parents, with their size and wing morphology remaining stable during the post-emigration phase [10, 11]. Consequently, variations in flyability of the landscape for these birds predominantly reflect their improved mastery of soaring flight, rather than changes in their morphology."

Discussion:

Line 154: Could the increase in step length also be due to changes in search strategies with age? e.g. from more Brownian motion when scavenging to Levy search patterns when actively hunting?

This is a very good point and we tried to look for evidence of this transition in the tracking data. We explored the first passage time for two individuals with a radius of 50 km to see if there is a clear transition from a Brownian to a Levy motion. The patterns that emerge are inconclusive and seem to point to seasonality rather than a clear transition in foraging strategy (Author response image 1). We have modified our statement in the discussion about the change in preference of step lengths indicating improve flight ability, to clarify that it is speculative:

First passage times using a 50 km radius for two randomly selected individuals.

"Our findings also reveal that as the eagles aged, they adopted longer step lengths, which could indicate an increasing ability to sustain longer uninterrupted flight bouts."

Methods:

Line 229: What is the cutoff for high altitude or high speed?

We used the Expectation-maximization binary clustering (EMbC) method to identify commuting flights. The EmbC method does not use hard cutoffs to cluster the data. Each data point was assigned to the distribution to which it most likely belonged based on the final probabilities after multiple iterations of the algorithm. Author response image 2 shows the distribution of points that were either used or not used based on the EmbC classification.

Golden eagle tracking points were either retained (used) or discarded (not used) for further data analysis based on the EmbC algorithm.

The point were clustered based on ground speed and height above ground.

Figure 1: The figure captions should stand on their own but in this case there is no information as to what the tests are actually showing.

We have now updated the caption to provide information about the model:

"Coefficient estimates of the step selection function predicting probability of use as a function of uplift proxies, week since emigration, and step length. All variables were z-transformed prior to modeling.

The error bars show 95% confidence intervals."

Reviewer 2 (Recommendations For The Authors):

First, I want to congratulate you on this fantastic work. I enjoyed reading it. The manuscript is clear and well-written, and the findings are sound and relevant to the field of movement ecology. Also, the figures are neatly presented and easy to follow.

I particularly liked expanding the old concept of fundamental vs realized niche into a movement ecology context. I believe that adds a fresh view into these widely accepted ecological assumptions on species niche, which may help other researchers build upon them to better understand movement "realms" on highly mobile animals in a rapidly changing world.

I made some minor comments to the manuscript since it was hard to find important weaknesses in it, given the quality of your work. However, there was a point in the discussion that I feel deserves your attention (or rather a reflection) on how major biological events such as moulting could also influence birds to master the flying and exploitation of the energy landscape. You may find my suggestion quite subjective, but I think it may help expand your idea for future works and, what is more, link concepts such as energy landscapes, ontogeny, and important life cycle events such as moulting in large soaring birds. I consider this relevant from a mechanistic perspective to understand better how individuals negotiate all three concepts to thrive and persist in changing environments and to maximise their

fitness.

Once again, congratulations on this excellent piece of research.

We thank the reviewer for their enthusiasm about our work and for bringing up important points about the biology of the species. Our detailed response are below.

MINOR COMMENTS:

(Note: Line numbers refer to those in the PDF version provided by the journal).

Line 110: Distinguished (?)

corrected

Line 131: Overall, I agree with the authors’ discussion and very much liked how they addressed crucial points. However, I have a point about some missing non-discussed aspects of bird ecology that had not been mentioned.

The authors argue that morphological traits are less important in explaining birds’ mastery of flight (thus exploiting all available options in the landscape). However, I think the authors are missing some fundamental aspects of bird biology that are known to affect birds’ flying skills, such as moult.

The moulting process affects species’ flying capacity. Although previous works have not assessed moults’ impact on movement capacity, I think it is worth including the influence of flyability on this ecologically relevant process.

For instance, golden eagles change their juvenile plumage to intermediate, sub-adult plumage in two or three moult cycles. During this process, the moulting process is incomplete and affects the birds’ aerodynamics, flying capacity, and performance (see Tomotani et al. 2018; Hedenström 2023). Thus, one could expect this process to be somewhat indirectly linked to the extent to which birds can exploit available resources.

Hedenström, A. (2023). Effects of wing damage and moult gaps on vertebrate flight performance.

Journal of Experimental Biology, 226(9), jeb227355. Tomotani, B. M., Muijres, F. T., Koelman, J., Casagrande, S., & Visser, M. E. (2018). Simulated moult reduces flight performance, but overlap with breeding does not affect breeding success in a longdistance migrant. Functional Ecology, 32(2), 389-401.

We thank the reviewer for bringing up this relevant topic. We explored the literature listed by the reviewer and also other sources. We came to the conclusion that moulting does not impact our findings. In our study, we included data for eagles that had emigrated from the natal territories, with their fully grown feathers in juvenile plumage. The moulting schedule in juvenile birds is similar to that of adults: the timing, intensity, and sequence of feathers being replaced is consistent every year (Author response image 3). For these reasons, we do not believe that moulting stage noticeably impacts flight performance at the scale of our study (hourly flights). Fine details of soaring flight performance (aerodynamics within and between thermals) could differs during moulting of different primary and secondary feathers, but this is something that would occur every time the eagle replaces these feather and we do not expect it to be any different for juveniles. Such fine scale investigations are outside the scope of this study.

Moulting schedule of golden eagles [12]

Lines 181-182: I don’t think trophic transitions rely only on individual flying skill changes. Furthermore, despite its predominant role, scavenging does not mean it is the primary source of food acquisition in golden eagles. This also depends on prey availability, and scavenging is an auxiliary font of easy-to-catch food.

Scavenging implies detecting carcasses. Should this carcass appearance occur in highly rugged areas, the likelihood of detection also reduces notably. This is not to say that there are not more specialized carrion consumers, such as vultures, that may outcompete eagles in searching for such resources more

efficiently.

In summary, I don‘t think such transition relies only on flying skills but on other non-discussed factors such as knowledge accumulation of the area or even the presence of conspecifics.

Line 183: This is precisely what I meant with my earlier comment.

Thank you for the discussion on the interaction between flight development and foraging strategy. We explored the transition from scavenging to hunting above as a response to Reviewer 1, but did not find a clear transition. This is in line with your comment that the birds probably use both scavenging and hunting methods opportunistically.

Lines 193-195: I will locate this sentence somewhere in this paragraph. As it is now, it seems a bit out of context. It could be a better fit at the end of the first point in line 203.

Thank you for pointing out the issue with the flow. We have now added a transitional sentence before this one to improve the paragraph. The beginning of the conclusion now reads as follows, with the new sentence shown in boldface.

"Spatial maps serve as valuable tools in informing conservation and management strategies by showing the general distribution and movement patterns of animals. These tools are crucial for understanding how animals interact with their environment, including human-made structures. Within this context, energy landscapes play an important role in identifying potential areas of conflict between animals and anthropogenic infrastructures such as wind farms. The predictability of environmental factors that shape the energy landscape has facilitated the development of these conservation tools, which have been extrapolated to animals belonging to the same ecological guild traversing similar environments."

References

(1) Colwell, R. K. & Rangel, T. F. Hutchinson’s duality: The once and future niche. Proceedings of the National Academy of Sciences 106, 19651–19658. doi:10.1073/pnas.0901650106 (2009).

(2) Corbeau, A., Prudor, A., Kato, A. & Weimerskirch, H. Development of flight and foraging behaviour in a juvenile seabird with extreme soaring capacities. Journal of Animal Ecology 89, 20–28. doi:10.1111/1365-2656.13121 (2020).

(3) Fuster, J. M. Frontal lobe and cognitive development. Journal of neurocytology 31, 373–385.

doi:10.1023/A:1024190429920 (2002).

(4) Ramsaran, A. I., Schlichting, M. L. & Frankland, P. W. The ontogeny of memory persistence and specificity. Developmental Cognitive Neuroscience 36, 100591. doi:10.1016/j.dcn.2018.09.002 (2019).

(5) Williams, H. J., Duriez, O., Holton, M. D., Dell’Omo, G., Wilson, R. P. & Shepard, E. L. C. Vultures respond to challenges of near-ground thermal soaring by varying bank angle. Journal of Experimental Biology 221, jeb174995. doi:10.1242/jeb.174995 (Dec. 2018).

(6) Williams, H. J., King, A. J., Duriez, O., Börger, L. & Shepard, E. L. C. Social eavesdropping allows for a more risky gliding strategy by thermal-soaring birds. Journal of The Royal Society Interface 15, 20180578. doi:10.1098/rsif.2018.0578 (2018).

(7) Harel, R., Horvitz, N. & Nathan, R. Adult vultures outperform juveniles in challenging thermal soaring conditions. Scientific reports 6, 27865. doi:10.1038/srep27865 (2016).

(8) Ruaux, G., Lumineau, S. & de Margerie, E. The development of flight behaviours in birds. Proceedings of the Royal Society B: Biological Sciences 287, 20200668. doi:10.1098/rspb.2020.

0668 (2020).

(9) Bolker, B., Warnes, G. R. & Lumley, T. Package gtools. R Package "gtools" version 3.9.4 (2022).

(10) Bortolotti, G. R. Age and sex size variation in Golden Eagles. Journal of Field Ornithology 55,

54–66 (1984).

(11) Katzner, T. E., Kochert, M. N., Steenhof, K., McIntyre, C. L., Craig, E. H. & Miller, T. A. Birds of the World (eds Rodewald, P. G. & Keeney, B. K.) chap. Golden Eagle (Aquila chrysaetos), version 2.0. doi:10.2173/bow.goleag.02 (Cornell Lab of Ornithology, Ithaca, NY, USA, 2020).

(12) Bloom, P. H. & Clark, W. S. Molt and sequence of plumages of Golden Eagles and a technique for in-hand ageing. North American Bird Bander 26, 2 (2001).

Version 2

Abstract

Abstract

The heterogeneity of the physical environment determines the cost of transport for animals, shaping their energy landscape. Animals respond to this energy landscape by adjusting their distribution and movement to maximize gains and reduce movement costs. Much of our current knowledge about energy landscape dynamics focuses on factors external to the animal, particularly the spatio-temporal variations of the environment. However, an animal’s internal state can significantly impact its ability to perceive and utilize the available energy, creating a distinction between the “fundamental” and the “realized” energy landscapes. Here we show that the realized energy landscape varies along the ontogenetic axis. Locomotor and cognitive capabilities of individuals change over time, especially during the early life stages. We investigate the development of the realized energy landscape in the Central European Alpine population of the golden eagle Aquila chrysaetos, a large predator that requires negotiating the atmospheric environment to achieve energy-efficient soaring flight. We quantified weekly energy landscapes using environmental features for 55 juvenile golden eagles, demonstrating that energetic costs of traversing the landscape decreased as the birds aged. In fact, the potentially flyable area within the Alpine region increased 2,170-fold during their first three years of independence. Our work contributes to achieving a predictive understanding of animal movement behaviors by presenting ontogeny as a mechanism shaping the realized energy landscape.

Introduction

The physical characteristics of the environment determine the energetic cost of transport for animals, defining their “energy landscape” [1, 2]. Consequently, temporal dynamics in environmental conditions can affect the inherently spatially-explicit energy landscapes across time. For example, patterns of thawing and freezing of lakes and rivers shape the migratory routes of Caribou, as walking on ice is energetically preferable to swimming [3]. Variations of the energy landscape along the temporal axis are most notable for animals that move in moving media: water or air. Sharks adjust their core habitat use to make the most of temporary updrafts formed by tidal currents colliding with upward-facing slopes, which help them reduce the energetic costs of remaining buoyant [4]. Temporal variations in atmospheric conditions shape hourly [2] and seasonally [5] optimal flight routes for birds.

The spatio-temporal variation in the energy landscape is predictable to a great extent. Animals reduce and thus optimize the cost of transport by tracking these changes as they move through the energy landscapes. As such, the distribution and movement patterns of animals can be largely explained by the energy landscape [6], making energy landscapes useful for understanding ecological and evolutionary processes. This knowledge is increasingly important for policy making to reduce human-wildlife conflicts, particularly for development of infrastructures [7, 8], and more generally in conservation planning [9]. Based on the timing and location of movement the energy landscape can be specific yet predictable for animals with similar morphology and movement modes, at the guild [2, 10], species [8], or sub-species [11] levels.

To date, our understanding of animals’ interaction with the energy landscape has depended on factors extrinsic to the moving animal. Reconstructing the energy landscape solely based on the characteristics of the environment can be conceptualized similar to Hutchinson’s concept of the fundamental ecological niche [12, 13]. Hutchinson defined the fundamental niche as the full range of environmental conditions in which a species can potentially exist and reproduce. Here we argue that the energy landscape too can, or possibly should, be conceptualized as the fundamental movement niche, defining the full range of physical environmental conditions within which a species can move while adhering to its movement-related energy budget, or sustainable cost of transport. However, just as animals do not occupy the entirety of their fundamental Hutchinsonian niche in reality [13], for example due to competition or predation risk, various factors can contribute to an animal not having access to the entirety of its fundamental movement niche. Our current understanding of the energy landscape lacks a differentiation between a fundamental and a realized energy landscape.

An animal’s realized energy landscape should depend on the motive to move and its internal state. For example, an animal’s decision-making when moving might be the outcome of its strategy to maximize long-term fitness rather than short-term energy savings. Halsey et al. [14] introduced the concept of “individual energy landscapes”, stating that the energy landscape fluctuates, for example, when an animal increases its speed to out-run a predator [14] or when infection or illness reduces an animal’s movement capacity [15, 16]. As a consequence, the realized energy landscape varies in time from the perspective of the moving animal, sometimes irrespective of, but in interaction with and as a consequence of, the immediate environment.

We would like to argue that the animals’ ability to exploit the energy landscape is transformed across one of the most fundamental progressions in all animals’ lives: the ontogenetic axis. In addition to morphological changes, as young animals progress through their developmental stages, their movement proficiency [17] and cognitive capabilities [18] improve and memory manifests [19]. Cognitive capabilities allow the animal to improve the perception of its environment and to make optimal decisions when choosing energy-efficient paths [20, 21]. Improved motor skills enable the animal to respond more appropriately to the perceived environment to utilize the maximum available energy and/or to avoid areas expensive to traverse [22–25]. And finally, memory allows the animal to recall past combinations of space and environmental conditions that affected its costs of movement [26]. Developmental changes and improvements in all these elements should be reflected in the animal’s re-alized energy landscape, where hotspots of energy gain within the landscape expand as young animals gain experience and master their specific movement behaviors.

We investigated the demographic shift in the realized energy landscape through gaining experience in a movement specialist, the golden eagle Aquila chrysaetos. As obligate soaring birds, golden eagles utilize ascending air currents, known as uplifts, to gain vertical altitude, which subsequently enables them to glide for horizontal displacement, without relying on energetically expensive wing flapping. Soaring flight is a learned and acquired behavior [24, 27], requiring advanced cognitive skills to locate uplifts as well as fine-tuned locomotor skills for optimal adjustment of the body and wings to extract the most energy from them, for example by adopting high bank angles in narrow and weak thermals [28] and reducing gliding airspeed when the next thermal has not been detected [29]. This flight mode enables soaring birds to cover long distances with minimal energy investments, almost as low as when resting [30]. We approach the energy landscape concept with a focus on energy availability. This is commonly done for soaring birds, as uplift and wind support can directly provide energy for flying birds [11, 31].

We analyzed 46,000 hours of flight data collected from bio-logging devices attached to 55 wild-ranging golden eagles in the Central European Alps. These data covered the transience phase of natal dispersal (hereafter post-emigration). In this population, juveniles typically achieve independence by emigrating from the parental territory within 4-10 months after fledging. However, due to the high density of eagles and consequently the scarcity of available territories, the transience phase between emigration and settling by eventually winning over a territory is exceptionally long at well over 4 years. Our hypothesis posited that the realized energy landscape during this transience phase gradually expands as the birds age. More specifically, we expected the habitat to become progressively cheaper to traverse as the birds aged, resulting in an expansion of flyable areas within the Alpine region as a consequence of gaining experience and flight proficiency.

Results

We explored the movement decisions of 55 juvenile golden eagles in the Central European Alps during their post-emigration commuting flights (i.e. non-stop flight bouts lasting at least 1 hour) from the first week until three years after emigration. Following a step-selection approach [32], at each movement step, we compared the observed step to 50 alternative steps that were available to each bird in space and time. These steps were compared in respect to the topography of the terrain, specifically Topographic Ruggedness Index (TRI) and distance to ridge lines, both useful proxies for occurrence of uplifts [31, 33]. The resulting step-selection model distinguishing between the used and available conditions with a good predictive performance based on the normalized Root Mean Squared Error (RMSE = 0.14). Overall, the birds preferred to fly over areas with high potential for uplift formation, characterized by high values of TRI and lower distance to ridge lines (Fig. 1). The birds also preferred to fly with long step lengths, a trend that increased with age. As the birds grew older, they were less likely to avoid areas with lower TRI and higher distance to ridge lines (Fig. 2). We found individual variation in the coefficients for TRI and distance to ridge line (Fig. S1).

Coefficient estimates of the step selection function predicting probability of use as a function of uplift proxies, week since emigration, and step length. All variables were z-transformed prior to modeling. The error bars show 95% confidence intervals.

Flyability index predicted using the step-selection model for combinations of topography and week since emigration values. The interactions between Topographic Ruggedness Index (TRI) and distance to ridge lines with week are among the set of criteria that young eagles used for selecting where to fly during their commuting flights.

We used the step-selection model to predict the flyability of the Alpine region for each developmental stage (week since emigration). This flyability index quantified the potential energy that a bird could obtain from its surroundings by considering topography and the bird’s movement. This index serves as a measure of the suitability of a location for energy-efficient flight, reflecting the bird’s ability to exploit favorable uplift conditions. We constructed energy landscapes for the juvenile golden eagles for each week since independence by predicting the flyability within the Alpine region. Based on these predictions, we determined the flyable area for the juvenile and immature golden eagles during the developmental phase at weekly increments. We defined flyable areas as the cells within a 100*100 m grid with a flyability value larger than 0.7. Our analysis show that hotspots of energy availability in the birds’ landscape expanded over time (Fig. 3; Video S1). We found that the flyable area increased 2,170-fold from the first week until three years after emigration. From an initial 0.038 % in the first week after emigration, the flyable area followed a logistic growth curve to plateau at 81 % of the entire Alpine region becoming flyable in the third year (Fig. 4). The tracking data indicated that the golden eagles flew over an area of 92,000 km2 by the third year after emigration (Fig. S2).

Hotspots of energy availability for golden eagles’ flight in the Alps. Flyable areas were defined as cells within a 100*100 m grid with predicted flyability above 0.7 based on our step-selection model. The maps show the 2D kernel density estimation of flyable areas for golden eagles at different timestamps since dispersal: week 1, week 4 (one month), week 24 (six months) and week 52 (one year). The raw prediction maps for every week since dispersal are shown in the video S1.

The flyable area for juvenile golden eagles in the Alpine region from the first week until three years after emigration. Flyable area was defined as the total number of cells within a 100*100 m grid with predicted flyability larger than 0.7 based on the step-selection model. The positive trend shows that juvenile golden eagles can fly over a larger portion of the Alpine region as they age.

Discussion

We show that the flyable area for juvenile golden eagles increased during the developmental period. This increase, we argue, is most likely linked to the birds’ improving ability to negotiate their atmospheric environment better as they aged, due to an increasing capacity to perceive and exploit the energy available in the landscape. Because we estimated flyability based on proxies of uplift formation, an important component shaping the energy availability landscape for soaring birds [11, 31], the gradual increase in flyable areas can be interpreted as a change in the realized energy landscape.

Our findings assert the concept of “individual energy landscapes”. Initially described by Halsey [14], this concept suggests that the energy expenditure of an animal when moving is more complex than the cost of transport as determined solely by its morphology and the external environment. The locomotor decisions that an animal makes to maximize its fitness at any given point in time would also affect its energy expenditure. We present evidence that ontogeny also serves as a potential mechanism giving rise to individual energy landscapes, not only due to its impact on decision-making, but also due to the influence it exerts on the available options for the animal. During the early stages of their development, animals may not possess the full capacity to perceive their environment, as they will in later stages [20, 21]. Additionally, their motor skills are still being honed [24]. Consequently, younger individuals may respond less efficiently to the environmental conditions they must navigate, in contrast to older, more proficient individuals.

By relying on static topographic variables to build the energy landscapes, we ensured that the environment remained constant over time. Juvenile golden eagles complete their morphological development before gaining independence from their parents, with their size and wing morphology remaining stable during the post-emigration phase [34, 35]. Consequently, variations in flyability of the landscape for these birds predominantly reflect their improved mastery of soaring flight, rather than changes in their morphology. Our findings also reveal that as the eagles aged, they adopted longer step lengths, which could indicate an increasing ability to sustain longer uninterrupted flight bouts.

Furthermore, our results provide insight into the development of different flight modes. These birds can switch between orographic and thermal soaring [36]. Orographic soaring refers to utilization of upward air currents generated when wind encounters and flows over elevated topographic features such as hills, mountains, or cliffs. Locating orographic uplifts requires less experience, as they form predictably at mountain ridges [37, 38], compared to locating and using more transient thermals which form less predictably over flat surfaces [31]. Our findings suggest that with increasing age, golden eagles place decreasing importance on ruggedness and proximity to ridge lines when deciding where to fly, indicating a learning progression from mastering orographic uplifts before tackling the more challenging thermal soaring.

The energy landscapes maps showed the amount of energy potentially available to juvenile golden eagles during progressing weeks since emigration (Video S1). These maps are based on a step-selection model built on the birds’ empirical use of the landscape, the energy availability thus represents the energy that would be theoretically perceivable and exploitable by the birds in a specific developmental phase or age. The detailed perspective on an energy landscape which transforms over time with the accumulation of experience highlights the process of changes in the realized movement niche that would be available to the animals at a certain age. These changes in the ability to extract energy from the atmospheric column will induce age-specific consequences on movement capacity and the cost of transport.

We argue that the realized energy landscape does not represent the realized ecological niche nor the distribution range of animals, but rather the subset of the fundamental energy landscape that is available to the individual based on its ability to perceive and exploit the energy in the landscape. In our study, the realized movement patterns of the golden eagles within the Alpine region covered a small portion of the area predicted as flyable. This is because the birds’ movement (and thus their realized individual and age-specific movement niches) will be defined not only by the energy landscape, but also by the interplay of philopatry, developmental requirements, the social landscape, and anthropogenic features. As golden eagles age, other crucial behaviors such as transitioning from predominantly scavenging to active hunting are also happening [39]. A transition that in itself also might depend on advanced flying skills. Thus, the choice of where to fly will not solely follow an energy-minimizing strategy while moving, but naturally also increasingly accommodate resource use opportunities. Additionally, older transient golden eagles are more susceptible to potentially deadly conflicts with territorial adult birds, influencing their movement decisions as they avoid established territories [40] affecting their landscape of fear and thus their decision landscape overall [41–43]. The distribution of anthropogenic features could also impact the eagles’ decision-making on where to fly [44]. The strong individual variation that we observed in the birds’ response to distance to ridge lines (Fig. S1) could indicate variable landscape utilization strategies among the juvenile golden eagles in response to these factors and their interaction.

Conclusion

Spatial maps serve as valuable tools in informing conservation and management strategies by showing the general distribution and movement patterns of animals. These tools are crucial for understanding how animals interact with their environment, including human-made structures. Within this context, energy landscapes play an important role in identifying potential areas of conflict between animals and anthropogenic infrastructures such as wind farms. The predictability of environmental factors that shape the energy landscape has facilitated the development of these conservation tools, which have been extrapolated to animals belonging to the same ecological guild traversing similar environments. However, with recent development of concepts of species-specific [8] and individual-specific energy landscapes [14], the realistic prediction of animals’ energy landscapes appears a multifaceted challenge. By identifying the distinct dimensions along which individual energy landscapes can vary, particularly along the demographic axis, we enhance our capacity to understand and predict these landscapes. Our research contributes to this broader understanding by revealing the stark influence of ontogeny in defining the aging of the realized energy landscape. By taking inspiration from the ecological niche definitions to conceptualize the energy landscape, we hope to improve further development of this concept, aiding the interpretation of variations in the energy landscape across many axes, both external and internal to the moving animal.

Materials and Methods

Bio-logging dataset

We used Global Positioning System tracking data from a long-term study of juvenile golden eagles in the resident population of Central European Alps. We used data collected for a total of 55 juveniles tracked over 2017-2023 (Table S1) with a regular sampling schedule of one GPS fix every 20 minutes during daytime (see [45] for details on fieldwork procedure and data sampling schedules). Nestlings were fitted with solar powered loggers manufactured by e-obs GmbH (Munich, Germany) using a leg-loop harness. The weight of the harness and logger (in total maximum 60g) was below the recommended maximum 3% of the birds’ body mass. We subset the data to correspond to the transient phase of natal dispersal up to three years after emigration. We identified emigration timing using recursions to the natal territory [46]. If an animal spent more than 14 consecutive days without recursions to a 7 km radius around its nest, we classified it as dispersed. These thresholds were chosen based on expert knowledge of golden eagle behavior in the region. For two animals that had exceptionally large natal territories to start with, we increased the radii to 30 km to compensate for this potential bias.

Track segmentation

We focused on commuting flights for investigating the energy landscape for the golden eagles. To extract commuting flights, we segmented the tracking data based on ground speed and flight altitude using Expectation-Maximization binary Clustering (EMbC) [47] applied at the population level (i.e. to all individuals). To calculate flight altitude above ground for each tracking point, we first calculated altitude above mean sea level as the difference between altitude above ellipsoid recorded by the bio-logging devices and the Global Geoid Model EGM96 (WGS 84). We then subtracted the ground elevation (Copernicus Digital Elevation Model; 25 m resolution https://land.copernicus.eu/) from altitude above mean sea level as an estimate of flight altitude. We labeled segments with high flight altitude and high speed as commuting flights and used these for further analysis. We sub-sampled the tracking data to hourly intervals to focus on commuting flights that were longer than one hour.

Topographic variables

We used two topographic variables as proxies for uplift availability: Topographic Ruggedness Index (TRI) and distance to ridge lines. The former describes the topographic heterogeneity of the terrain by calculating the mean of the absolute elevation difference between a cell and its adjacent cells [48] and the latter represents a proxy for orographic updrafts generated by deflection of the air current by a physical barrier [33, 37]. To extract the ridge lines and TRI values, we used the Copernicus Digital Elevation Model. We identified the ridge lines based on the Weisss landforms classification using Topographic Position Index (TPI), a scale-dependent index which compares the elevation of each cell to the neighboring cells within a given radius [49]. As the landform categories depend on the size of the radius, several TPI classifications at different scales were obtained using different radii [50, 51]. We selected a classification using a 10 m inner radius and 200 m outer radius on the basis of the narrow size of the identified ridge lines. Then, we calculated the distance from each pixel to the ridge line pixel using the function proximity raster from SAGA [52] to obtain a distance map to the ridge line. For both TRI and distance to ridge lines, we averaged neighboring cell values to achieve a 100 m cell size.

Modeling the energy landscape

We used a step-selection approach [32] to summarize the environmental conditions in which the birds chose to commute. Each pair of consecutive points in the tracking dataset was considered as one step. For each step, we generated 50 random steps that originated at the observed steps’ start point, but had alternative end points to represent alternative movement decisions that the bird could have taken. The location of the alternative end points were selected by drawing from a gamma distribution fitted to the step lengths and a von Mises distribution fitted to the turning angles of all the empirical tracking data. The result was a stratified dataset with one used and 50 alternative points in each stratum.

All used and available points were associated with TRI and distance to ridge line values. We found no autocorrelation (r < 0.5) among our predictor variables TRI, distance to ridge line, step length, and week since emigration (as a continuous variable). The predictor variables were z-transformed prior to modeling to ensure comparability of the coefficient estimates. We used a conditional logistic regression to build a step-selection function to model and predict the energy landscape. We included step length as one of the predictors in our model, to take the movement characteristics of the growing juveniles into account [32]. To track the changes in the importance of the topographic variables over time, we included three-way interaction terms for the TRI and distance to ridge line, with step length and week since emigration. To account for individual variation, we included the random effect of individuals on the slopes of TRI and distance to ridge line. The model was fit using the Template Model Builder (TMB; [53]) in R [54], following the procedure suggested by Muff et al. [55].

To test the predictive performance of the model, we computed the normalized Root Mean Squared Error (RMSE) [56]. This metric can be interpreted as the standard deviation of the unexplained variance, with lower values indicating a smaller difference between observed data and the model’s predicted values.

To better interpret the interaction terms between the topographic variables and week since emigration, we made predictions using the model for different combinations of TRI and week as well as distance to ridge line and week. In the new datasets that we generated to make these predictions, we set the variables that were not represented in the interaction term to their average values and set the grouping factor that represented the stratum as missing data. We made the predictions on the scale of the link function and converted them to values between 0 and 1 using the inverse logit function [57]. These predicted values estimated the probability of use of an area based on the model. We interpreted these predicted values as the flyability index, representing the potential energy available in the landscape to support flight, based on the uplift proxies (TRI and distance to ridge line) and the movement capacity (step length) of the birds included in the model.

Energy landscape maps

To investigate the changes in the energy landscape across the onto-genetic axis, we used our model to create prediction maps for weekly increments during the first three years after emigration. We made our prediction maps for the entire extent of the Alpine region (“Perimeter of the Alpine Convention” layer available via Permanent Secretariat of the Alpine Convention). For each week since emigration, we created a new dataset to make predictions for each 100 m x 100 m cell. We assigned the mean step length value for each given week in the new datasets.

To explore the changes in flyable areas over time, we defined landscape flyability as flyability values above 0.7. We then estimated the total flyable areas for each week since emigration. The Alpine region covers 190,544.7 km2 (based on the Alpine perimeter layer mentioned above). As a final step, we compared the eagles’ flyable area in weeks 1 and 156 with the total area of the Alpine region.

Supporting information